---
title: "Suns Project #2 Code & Notes"
author: "Joe Siwinski"
date: "9/11/2020"
output:
  html_document:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, error = FALSE, warning = FALSE, fig.align = "center", message = FALSE)
```

---

# Introduction

The goal of this project is to predict win totals and probability of making the playoffs for any team prior to the start of the season. In addition, we want to find out the impact on playoff probabilty for teams when their is a play-in series. I took the following approach to complete the project. These steps are expanded on in the sections that follow.

<br/>

<center>

*Outline of my steps to complete the project*

</center>

<br/>

1) Model #1 - Game Score Differential Model


2) Model #2 - Game Win Prob Model (Based on Predicted Score Differenctial)


3) Simulate 10,000 seasons on a "fake" 82 game traditional schedule to estimate average end of season win totals and playoff probability


4) Simulate a play-in series for seasons from 10,000 sims that meet the play-in criteria and compare playoff probabilities of play-in seasons to traditional seasons in step #3.

<br/>

---

## 1 - Score Differential Model

I will first train a model on each game in the past 5 seasons (2014-15 to 2018-19 Season) to come up with an expected score differential between two teams that are playing. Since we are working strictly on predictions made prior to the season starting, we cannot use any performance metrics that the team has (Net Rating, Offensive Net Rating, etc.). In other words, I am confined to using roster variables (Amount of stars on the team, Career Win Share avg. of team, etc.) and location of game (Home/Away) to predict this score differential. To keep things simple, I will train this model within relation to the home team only. The resulting underlying model function is shown below.


<br/>

<center>

$$\text{Home Team Score Diff}  =  \beta{0}+ \beta1*{\text{Roster OWS Diff.}} + \beta2*{\text{Roster DWS Diff.}} +  \beta3*{\text{Roster All-NBA Diff.}} + \epsilon $$

</center>

<br/>

I will gather data for each team for the 2014-15 to 2018-19 NBA seasons to train this model. For Win Share values, I will use each players' previous 2 season Win Share averages. For All NBA, I will utilize the past seasons All-NBA selections. This data gather is shown in the code chunk below.

```{r, eval=FALSE, echo=TRUE}

# Load packages
library(tidyverse)
library(ggplot2)
library(glue)
library(scales)
library(plotly)
library(shiny)
library(readxl)

########## NEED DATA FOR TRAINING ###########

# Read in game roster data 
game_rosters <- read_rds("../SunsProject/game_data.rds")

# Read in All NBA team selections
all_nba <- read_rds("../SunsProject/all_nba_selections.rds")

# Read in game results data
game_results <- read_rds("../SunsProject/game_results.rds")

library(ballr)

# Trying to get season stats from past two seasons for each season start
fetch_winshares <- function(x){
  
  # Getting Per game OWS DWS
  test <- NBAPerGameAdvStatistics(x) %>% 
    mutate(tot_mins = g*mp) %>% 
    select(player, tot_mins, ows, dws)
  
  # Adding season 
  test$season <- x
  
  print(x)
  return(test)
  
}

# Loop through to get player win shares for each season since 2012-13
win_shares <- lapply(c(2013:2020), fetch_winshares)
win_shares <- do.call(rbind, win_shares)

# Getting TOT rows for traded players
win_shares <- win_shares %>% 
  group_by(player, season) %>% 
  filter(tot_mins == max(tot_mins))

# Getting previous 2 season weighted average win shares for each season
library(zoo)
win_shares <- win_shares %>% group_by(player) %>% arrange(desc(season)) %>%
  mutate(tot_mins_sum = rollsum(tot_mins, k = 2, fill = dplyr::lead(tot_mins)),
         
         
         # Getting OWS/DWS * tot_mins for each season
         ows_mins = ows*tot_mins, 
         dws_mins = dws*tot_mins,
         
         # Getting OWS/DWS * tot_mins sum for past two seasons
         ows_sum = rollsum(ows_mins, k = 2, fill =  dplyr::lead(ows_mins)), 
         dws_sum = rollsum(dws_mins, k = 2, fill =  dplyr::lead(dws_mins)), 
         
         # Divide OWS/DWS sums by tot_mins_sum column to get weighted average
         ows_wa = ows_sum/tot_mins_sum, 
         dws_wa = dws_sum/tot_mins_sum) %>% 
  
  # Getting vars needed
  select(player, season, ows_wa, dws_wa, tot_mins = tot_mins_sum) %>% 
  mutate(season = season + 1) %>% 
  
  ungroup()


# Putting in 25th percentile avg. win shares for rookies
ows_missing <- quantile(win_shares$ows_wa, na.rm = TRUE, probs = 0.25) %>% 
  as.numeric()
dws_missing <- quantile(win_shares$dws_wa, na.rm = TRUE, probs = 0.25) %>% 
  as.numeric()

win_shares <- win_shares %>% mutate(ows_wa = ifelse(is.na(ows_wa) == TRUE, ows_missing, ows_wa), 
                      dws_wa = ifelse(is.na(dws_wa) == TRUE, dws_missing, dws_wa))



# Getting rid of accents in player names
rm_accent <- function(str,pattern="all") {
   if(!is.character(str))
    str <- as.character(str)

  pattern <- unique(pattern)

  if(any(pattern=="Ç"))
    pattern[pattern=="Ç"] <- "ç"

  symbols <- c(
    acute = "áéíóúÁÉÍÓÚýÝ",
    grave = "àèìòùÀÈÌÒÙ",
    circunflex = "âêîôûÂÊÎÔÛ",
    tilde = "ãõÃÕñÑ",
    umlaut = "äëïöüÄËÏÖÜÿ",
    cedil = "çÇćčşšŽžūİŠāŠ"
  )

  nudeSymbols <- c(
    acute = "aeiouAEIOUyY",
    grave = "aeiouAEIOU",
    circunflex = "aeiouAEIOU",
    tilde = "aoAOnN",
    umlaut = "aeiouAEIOUy",
    cedil = "cCccssZzuISaS"
  )

  accentTypes <- c("´","`","^","~","¨","ç")

  if(any(c("all","al","a","todos","t","to","tod","todo")%in%pattern)) # opcao retirar todos
    return(chartr(paste(symbols, collapse=""), paste(nudeSymbols, collapse=""), str))

  for(i in which(accentTypes%in%pattern))
    str <- chartr(symbols[i],nudeSymbols[i], str) 

  return(str)
}

# Joining the 2 year avg. dws ows to the game_rosters df
game_rosters2 <- game_rosters %>% 
  
  
  # Joining all - NBA Selection players
  left_join(all_nba %>% dplyr::select(player, season, AllNBA)) %>%
  
  # Getting rid of Jr. and Sr. 
  mutate(player = str_remove_all(player, '\\ Jr.'), 
         player = str_remove_all(player, '\\ Sr.')) %>%
  
  left_join(win_shares %>%
  ungroup() %>%
  mutate(player = str_replace(player, '\\*', ''), 
         player = str_remove_all(player, '\\.'), 
         player = str_remove_all(player, '\\ Jr'), 
         player = str_remove_all(player, '\\ Sr'), 
         player = trimws(player)) %>%
  mutate(player = rm_accent(player)), by = c("player" = "player", "season" = "season")) %>% 
  filter(is.na(dws_wa) == FALSE) 

### GO BACK TO THIS TO ADJUST ROOKS


# Setting up model dataframe 
# Need dws, ows, amount of ALL NBA for each team game
team_ws_game <- game_rosters2 %>% filter(min > 0) %>% 
  filter(grepl('Regular Season', dataset)) %>%
  select(season, gameid, date, player, own_team, opp_team, venue, ows_wa, dws_wa, AllNBA) %>% 
  
  #Getting sum of ows_wa
  group_by(gameid, season, date, own_team, opp_team, venue) %>%
  summarise(tot_ows = sum(ows_wa), 
            tot_dws = sum(dws_wa), 
            tot_all_nba = sum(AllNBA, na.rm = TRUE)) 


# Each game has one row in order to obtain differences in each stat with respect to home team
home_team_games <- team_ws_game %>% 
  ungroup() %>%
  dplyr::rename(home_ows = tot_ows, home_dws = tot_dws, home_all_nba = tot_all_nba) %>%
  filter(venue == "H") %>% 
  left_join(team_ws_game %>% ungroup() %>% filter(venue == "R") %>% dplyr::select(gameid, season, away_team = own_team,
                                                             away_ows = tot_ows, away_dws = tot_dws, away_all_nba = tot_all_nba), 
            by = c("gameid", "season", "opp_team" = "away_team")) %>% 
  
  # Join game scores 
  left_join(game_results %>% 
              select(gameid, home_final_score, away_final_score = road_final_score, season), 
            by = c("gameid", "season")) %>% 
  
  
  mutate(ows_diff = home_ows - away_ows, 
         dws_diff = home_dws - away_dws, 
         all_nba_diff = home_all_nba - away_all_nba, 
         score_diff = home_final_score - away_final_score) %>% 
  
  select(gameid, season, ows_diff:score_diff)


### MODEL DF NOW READY - home_team_games ##

write_rds(home_team_games, "winprob_model_data.rds")

```

---

### Variable Relationships

Once I collect the data into one condense data frame, I am now ready to create my first model that predicts the *score differential* with respect to the home team. We can first explore the three variables and their relationships with the dependent variable (Score Differential) in the plots below:


```{r, fig.align='center', fig.height=8, echo=TRUE}
# Packages
library(tidyverse)
library(glue)
library(ggplot2)
library(ggpubr)


# Read in data
home_team_games <- read_rds("../SunsProject/winprob_model_data.rds") 


### Variable exploration ###

# Plot maker function
plot_maker <- function(stat_selected = c("ows_diff", "dws_diff", "all_nba_diff"), title_selected) {
  
  
  p <- home_team_games %>%
  ggplot(aes(x = home_team_games[[stat_selected]], y = score_diff)) + 
  geom_point() + 
  geom_smooth(method = "glm") + 
  stat_cor() + 
  
    
  # Theme stuff
  theme_classic() + 
  ggtitle(glue("{title_selected}")) + 
  xlab(glue("{title_selected}")) + 
  ylab("Score Differential") +
  theme(plot.title = element_text(hjust = 0.5))

  
}

#Running function for each stat diff.
p1 <- plot_maker("ows_diff", "OWS Diff.")
p2 <- plot_maker("dws_diff", "DWS Diff.")
p3 <- plot_maker("all_nba_diff", "All-NBA Diff.")


# Output to Rmd grid.arrange
gridExtra::grid.arrange(p1,p2,p3)

write_rds(p1, "~/Documents/Bookdown Project 2021/p1.rds")
write_rds(p2, "~/Documents/Bookdown Project 2021/p2.rds")
write_rds(p3, "~/Documents/Bookdown Project 2021/p3.rds")

```


---

### Score Difference Model Training

We have encouraging variable correlations as shown in the previous plots. Now it is time to train a model to predict score differential of games. I am going to train a model using `cv.glmnet` from the `glmnet` package in R. Again, this model is trained on the following underlying function:

<br/>

<center>

$$\text{Home Team Score Diff}  =  \beta{0}+ \beta1*{\text{Roster OWS Diff.}} + \beta2*{\text{Roster DWS Diff.}} +  \beta3*{\text{Roster All-NBA Diff.}} + \epsilon $$

</center>

<br/>


```{r, echo=TRUE, eval=FALSE}

# Data to be used here is again home_team_games.rds

#############################
###### MODEL TRAINING #######
#############################

# Packages
library(tidyverse)
library(broom)
library(glmnet)
library(coefplot)

# Output vars
y <- home_team_games$score_diff
x <- home_team_games %>% select(ows_diff, dws_diff, all_nba_diff) %>% 
  
  # Scale for interprettable coef estimates
  # mutate_all(scale) %>%
  data.matrix()

lambdas <- 10^seq(3, -2, by = -.1)


# Tuning for lambda
cv_fit <- cv.glmnet(x, y, alpha = 0, lambda = lambdas)
plot(cv_fit)

# Minimized lambda
opt_lambda <- cv_fit$lambda.min
opt_lambda

# Fit
fit <- cv_fit$glmnet.fit
summary(fit)
coef(cv_fit, "lambda.min")

# Predict 
y_predicted <- predict(fit, s = opt_lambda, newx = x)

# Sum of Squares Total and Error
sst <- sum((y - mean(y))^2)
sse <- sum((y_predicted - y)^2)

# R squared
rsq <- 1 - sse / sst
rsq


# Prediction dataframe
home_team_games_preds <- cbind(home_team_games, y_predicted) %>% 
  dplyr::rename(pred_score_diff = `1`)
```

---

## 2 - Win Prob Model

After training a model to predict score differential, we can then train a model that predicts win probability based on the score prediction. To accomplish this, I utlized a logistic glm modeling technique. The formula that the win-prob model is based on is as follows:

<br/>

<center>

$$\text{Win Probability}  =  \beta{0}+ \beta1*{\text{Predicted Score Diff.}} + \epsilon $$

</center>

<br/>


```{r, eval=FALSE, echo=TRUE}

# Utilizing created dataframe in previous code chunk - home_team_games_preds
# Creating a win binary column
home_team_games_preds <- home_team_games_preds %>% 
  mutate(win = ifelse(score_diff > 0, 1, 0))

# Model function will be:
# Prob Winning Game ~ Predicted Score Differential 

# Train/Test Split
require(caTools)
set.seed(824) 
sample_split <- sample.split(home_team_games_preds$win, SplitRatio = .75)
train = subset(home_team_games_preds, sample_split == TRUE)
test  = subset(home_team_games_preds, sample_split == FALSE)

# Inputs
y <- train$win
x <- train$pred_score_diff

# GLM 
model_winprob <- glm(win ~ pred_score_diff, family = "binomial", data = home_team_games_preds)

# Predict on both training and test sets
train <- train %>% mutate(win_prob = predict(model_winprob,., type = "response")) 
test <- test %>% mutate(win_prob = predict(model_winprob, ., type = "response"))

# LogLoss Evals on both train and test sets
library(MLmetrics)
LogLoss(train$win_prob, train$win)
LogLoss(test$win_prob, test$win)
            
```

```{r, echo=FALSE, eval=FALSE}
# Train Calibration Plot
train_cal_plot <- train %>% 
  mutate(win_prob = plyr::round_any(win_prob, 0.1)) %>% 
  group_by(win_prob) %>% 
  summarise(tot = n(), 
            actual_win_perc = mean(win)) %>% 
  
  ggplot(aes(x = win_prob, y = actual_win_perc, size = tot, 
             text = glue("Predicted Win Prob. = {win_prob}
                         Actual Win Prob. = {round(actual_win_perc,2)}
                         Total Observations = {tot}"))) + 
  geom_point() + 
  geom_abline() + 
  
  # Theme stuff
  theme_classic() +
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5)) + 
  scale_x_continuous("Predicted Win Probability", limits = c(0,1)) + 
  scale_y_continuous("Actual Win Probaility", limits = c(0,1)) + 
  ggtitle("Train Set - Win Prob Calibration Plot")

# Test Calibration Plot
test_cal_plot <- test %>% 
  mutate(win_prob = plyr::round_any(win_prob, 0.1)) %>% 
  group_by(win_prob) %>% 
  summarise(tot = n(), 
            actual_win_perc = mean(win)) %>%
  
  ggplot(aes(x = win_prob, y = actual_win_perc, size = tot, 
             text = glue("Predicted Win Prob. = {win_prob}
                         Actual Win Prob. = {round(actual_win_perc,2)}
                         Total Observations = {tot}"))) + 
  geom_point() + 
  geom_abline() + 
  
  # Theme stuff
  theme_classic() +
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5)) + 
  scale_x_continuous("Predicted Win Probability", limits = c(0,1)) + 
  scale_y_continuous("Actual Win Probaility", limits = c(0,1)) + 
  ggtitle("Test Set - Win Prob Calibration Plot")
   
# Write to rds
write_rds(train_cal_plot, "../SunsProject/train_cal_plot.rds")
write_rds(test_cal_plot, "../SunsProject/test_cal_plot.rds")

```

To evaluate my win probability model, I also like to create calibration plots as shown below. Here we can see that the model does a good job of predicting game outcomes on both the train data as well as the test dataset.

<center>

*Calibration Plots on Win Prob Predictions*

</center>

```{r, echo=FALSE}
# Read in cal plot rds
train_cal_plot <- read_rds("../SunsProject/train_cal_plot.rds")
test_cal_plot <- read_rds("../SunsProject/test_cal_plot.rds")

# Output to Rmd
library(shiny)
library(plotly)
div(ggplotly(train_cal_plot, tooltip = c("text")), align = "right")
div(ggplotly(test_cal_plot, tooltip = c("text")), align = "right")
```

---

## 3 - Simulating Traditional Seasons {.tabset}

With a pre-season game win-prob model in place, we now theoretically know the probability of the home team winning the game in any matchup between two teams. For my simulations, I used a "fake" schedule by using the 2013-14 NBA season (not included in modeling process). Each game has a win probability associated with it. I simulated this "fake" season 10,000 times to obtain both win total average predictions and average playoff probability for each team based on all of the simulation results. 

First, I needed to gather the variable data for this past season (2019-20 Season) in order to make score differential & win probability predictions. In the following chunk, I also create the fake schedule dataframe that I use to simulate multiple seasons.

```{r, eval=FALSE, echo=TRUE}

# Read in starting rosters 2019-20 NBA season
starting_rosters <- read_excel("../SunsProject/WinProb.xlsx", sheet = "Initial_Rosters") %>% 
  
  # Select Columns player, team, injury
  select(player, team, injury) %>%
  
  # Mutate 0s for non injuries
  mutate(injury = ifelse(is.na(injury) == TRUE, 0, injury), 
         season = as.numeric(2020), 
         player = rm_accent(player),
         player = trimws(player),
         player = as.character(player)) %>% 
  
  # Remove na players
  na.omit() %>% as.data.frame() %>% 
  
  # Remove Blake Griffin Injury
  filter(!(player %in% c("Blake Griffin")))

# Read in all-NBA from previous season 
all_nba_twenty <- read_rds("../SunsProject/all_nba_selections.rds") %>% 
  filter(season == 2020)

## Getting win share averages for players at the start of the 2019-20 NBA season
team_data_for_sim <- starting_rosters %>% ungroup() %>%
  left_join(win_shares %>% 
              ungroup() %>%
              mutate(player = rm_accent(player), 
                     player = trimws(player))) %>% 
  arrange(team) %>% 
  
  # Rookie adjustment
  mutate(ows_wa = ifelse(is.na(ows_wa) == TRUE, ows_missing, ows_wa), 
         dws_wa = ifelse(is.na(dws_wa) == TRUE, dws_missing, dws_wa)) %>%
  
  # Porzingis adjustment
  mutate(ows_wa = ifelse(player == "Kristaps Porzingis", 2.0570608, ows_wa), 
         dws_wa = ifelse(player == "Kristaps Porzingis", 1.997118, dws_wa)
         ) %>% 
  
  select(-tot_mins) %>% 
  
  # Joining to get all NBA  
  left_join(all_nba_twenty %>% select(player, season, AllNBA), by = c("player", "season")) %>% 
  mutate(AllNBA = ifelse(is.na(AllNBA) == TRUE, 0, AllNBA)) %>% 
  

  # Removing players that have season injuries (KD)
  filter(injury == 0) %>% 
  unique() %>%
  #  
  # View()
  # 
  # Grouping by team and getting sum for ows, dws, AllNBA
  group_by(team) %>% 
  summarise(ows = sum(ows_wa, na.rm = TRUE), 
            dws = sum(dws_wa, na.rm = TRUE), 
            AllNBA = sum(AllNBA, na.rm = TRUE), 
            tot = n())

# Abbreviation translator 
abbreviation <- team_data_for_sim %>% distinct(team) %>% 
  pull(team)
team_names <- c("Atlanta", "Brooklyn", "Boston", "Chicago", "Charlotte", "Cleveland", "Dallas", "Denver", "Detroit", 
                "Golden State", "Houston", "Indiana", "LA Clippers", "LA Lakers", "Memphis", "Miami", "Milwaukee", 
                "Minnesota", "New Orleans", "New York", "Oklahoma City", "Orlando", "Philadelphia", "Phoenix", "Portland",
                "Sacramento", "San Antonio", "Toronto", "Utah", "Washington")
team_translator <- cbind(abbreviation, team_names) %>% as.data.frame()

# Getting team name in team data df
team_data_for_sim <- team_data_for_sim %>% left_join(team_translator, by = c("team" = "abbreviation")) %>% 
  select(-team) %>% 
  select(team = team_names, ows, dws, AllNBA)

# Reading in fake 82-game schedule (2013-14 NBA Schedule)
sim_schedule <- read_excel("../SunsProject/SimulationSchedule.xlsx") %>% 
  
  # Getting only Reg Season
  dplyr::filter(`DATASET` == "2013-2014 Regular Season") %>%
  
  #Getting what I need
  select(date = DATE, home_team = `HOME TEAM`, away_team = `ROAD TEAM`) %>% 
  
  # Joining to get home team team stats
  left_join(team_data_for_sim %>% 
              
              # Rename vars for home
              dplyr::rename(home_ows = ows, home_dws = dws, home_all_nba = AllNBA), 
            by = c("home_team" = "team")) %>% 
  
  # Joining to get away team team stats
  left_join(team_data_for_sim %>% 
              
              # Rename vars for away
              dplyr::rename(away_ows = ows, away_dws = dws, away_all_nba = AllNBA), 
            by = c("away_team" = "team")) %>% 
  
  
  # Getting differentials for model preds.
  mutate(ows_diff = home_ows - away_ows, 
         dws_diff = home_dws - away_dws, 
         all_nba_diff = home_all_nba - away_all_nba) %>% 
  
  select(date, home_team, away_team, ows_diff:all_nba_diff)
  
  
################################
## SIM SCHEDULE NOW COMPLETED ##
################################

```

Next, I simulate 10,000 seasons in the traditional format (No Play-in). The code to this and the results from the simulations is shown below.

```{r, eval=FALSE, echo=TRUE}

#######################
### SETTING UP SIMS  ##
#######################

# West 
west <- c("Dallas", "Denver",
                "Golden State", "Houston", "LA Clippers", "LA Lakers", "Memphis",
                "Minnesota", "New Orleans", "Oklahoma City", "Phoenix", "Portland",
                "Sacramento", "San Antonio", "Utah")

# East
east <- c("Atlanta", "Brooklyn", "Boston", "Chicago", "Charlotte", "Cleveland", "Detroit", 
                "Indiana", "Miami", "Milwaukee", 
                 "New York", "Orlando", "Philadelphia", "Toronto", "Washington")


# Setting up Sim Empty Data frames that will have win total for season and if they make the playoffs
simulated_west <- data.frame(west) %>% mutate(sim_wins = NA, sim_playoffs = NA)
simulated_east <- data.frame(east) %>% mutate(sim_wins = NA, sim_playoffs = NA)

# Applying Models to Sim Data for predicted Score Diff. and Win Prob
sim_model_data <- sim_schedule %>% dplyr::select(ows_diff:all_nba_diff) %>% 
  # mutate_all(scale) %>% 
  data.matrix()

sim_preds <- predict(fit, s = cv_fit$lambda.min, newx = sim_model_data) %>% as.data.frame() %>% 
  dplyr::rename(pred_score_diff = `1`) %>% 
  mutate(win_prob = predict(model_winprob, ., type = "response"))

# Sim data with preds 
sim_data_final <- sim_schedule %>% cbind(sim_preds)

# Simulating
season_simulator <- function(iter) {
  
  #To simulate a season, we generate 1230 random numbers between 1 and 0, one for each game of the season
  sim_random <- runif(1230)
  
  # Join to sim_data_final
  sim_df <- sim_data_final %>% cbind(sim_random) %>% 
    mutate(sim_win = ifelse(win_prob >= sim_random, 1,0), 
           sim_season = iter)
  
  # Home Wins
  home_wins <- sim_df %>% group_by(home_team) %>% 
    summarise(home_wins = sum(sim_win))
  
  # Away Wins
  away_wins <- sim_df %>% group_by(away_team) %>% 
    filter(sim_win == 0) %>%
    summarise(away_wins = n())
  
  # Total Wins
  output_df <- home_wins %>% left_join(away_wins, by = c("home_team"="away_team")) %>% 
    mutate(total_wins = home_wins+away_wins) %>% 
    dplyr::select(team = home_team, total_wins) %>% 
    mutate(sim_season = iter)
  
  # Progress
  print(percent(iter/10000))

  
  # Return Sim_df
  return(output_df)
  
  
}

# Running on 10,000 simulations
sims <- lapply(c(1:10000), season_simulator)
sims <- do.call(rbind, sims) %>%
  
  # Getting conference 
  mutate(conf = ifelse(team %in% west, "West", "East"))

# TRADITIONAL SEASON RESULTS
# Estimated Win Totals and Playoff Prob without play in
traditional_results <- sims %>% group_by(sim_season, conf) %>% 
  arrange(desc(total_wins)) %>% 
  mutate(playoff_seed = rank(-total_wins, ties.method = "random"), 
         playoffs = ifelse(playoff_seed <= 8, 1, 0)) %>% 
  group_by(team, conf) %>%
  summarise(avg_wins = mean(total_wins, na.rm = TRUE), 
            playoff_prob = mean(playoffs))

# Write to rds
write_rds(traditional_results, "../SunsProject/sim_traditional_results.rds")
```



### East Sim Results


```{r, echo=FALSE}
library(DT)

# Read in data
traditional_results <- read_rds("../SunsProject/sim_traditional_results.rds") %>% 
  mutate(avg_wins = round(avg_wins, 2)) %>%
  rename(Team = team, Conference = conf, `Avg Wins` = avg_wins, `Playoff Prob.` = playoff_prob) %>% 
  arrange(desc(`Avg Wins`))

## DATATABLE OUTPUT ##

DT::datatable(traditional_results %>% filter(Conference == "East"), selection = "single",
                  filter = 'top',
                  options = list(sDom  = '<"top"><"bottom">', 
                                  pageLength = 15)) %>% 
      formatPercentage('Playoff Prob.') %>%
      formatStyle('Playoff Prob.', background = styleColorBar(traditional_results$`Playoff Prob.`, 'lightblue'))


```

---

### West Sim Results

```{r, echo=FALSE}
library(DT)

# Read in data
traditional_results <- read_rds("../SunsProject/sim_traditional_results.rds") %>% 
  mutate(avg_wins = round(avg_wins, 2)) %>%
  rename(Team = team, Conference = conf, `Avg Wins` = avg_wins, `Playoff Prob.` = playoff_prob) %>% 
  arrange(desc(`Avg Wins`))

## DATATABLE OUTPUT ##

DT::datatable(traditional_results %>% filter(Conference == "West"), selection = "single",
                  filter = 'top',
                  options = list(sDom  = '<"top"><"bottom">', 
                                  pageLength = 15)) %>% 
      formatPercentage('Playoff Prob.') %>%
      formatStyle('Playoff Prob.', background = styleColorBar(traditional_results$`Playoff Prob.`, 'lightblue'))


```


---

## 4 - Simulating Play-Ins

Finally, I needed to compare playoff probabilities in a traditional season versus that of play-in style seasons. To do this, I simulated a play-in series for the simulated seasons from above that would have resulted in play-in games (Seasons where the 8 and 9 seeds were separated by 2 GB). With the game level win prob model, I simulated a play in series for each of these seasons and recalculated teams' playoff probabilities only now with the results of play-ins included. The code process to do this can be seen below.

<center>

*NOTE: Play in series has 8 seed at home for 1st game. 9 seed at home for 2nd game if necessary.*

</center>


```{r, echo=TRUE, eval=FALSE}
######################
#### PLAY IN SIMS ####
######################


# Getting sim seasons from section #3 that would have resulted in play in
play_in_sims <-  sims %>% group_by(sim_season, conf) %>% 
  mutate(
    games_behind = max(total_wins) - total_wins,
    playoff_seed = rank(-total_wins, ties.method = "random")) %>% 
  arrange(desc(playoff_seed)) %>%
  mutate(
    
    # Flag for if a team is in play-in tourney
    play_in = ifelse(playoff_seed == 9 & abs(games_behind - dplyr::lead(games_behind)) <= 2, 1, 0)) %>% 
  
  mutate(
    play_in = ifelse(playoff_seed == 8 & sum(play_in) == 1, 1, play_in), 
    play_in = ifelse(is.na(play_in) == TRUE, 0, play_in)) 


# Getting the iter number for these play in seasons
play_in_seasons <- play_in_sims %>% 
  filter(play_in == 1) %>% ungroup() %>%
  distinct(sim_season) %>% pull(sim_season)


# Sim function to play out play ins
# x = sim season iter number
play_the_play_in <- function(x){
  
  # x <- 1
  
  # Filtering the sim season iter
play_in_sims <- sims %>% group_by(sim_season, conf) %>% 
  
  # Games behind and playoff seed
  mutate(
    games_behind = max(total_wins) - total_wins,
    playoff_seed = rank(-total_wins, ties.method = "random")) %>% 
  arrange(desc(playoff_seed)) %>%
  
  # Flag for if team is in play in
  mutate(play_in = ifelse(playoff_seed == 9 & abs(games_behind - dplyr::lead(games_behind)) <= 2, 1, 0), 
    play_in = ifelse(playoff_seed == 8 & sum(play_in) == 1, 1, play_in), 
    play_in = ifelse(is.na(play_in) == TRUE, 0, play_in)) %>% 
  
  # filter iter
  filter(sim_season == x)

# filtering 8-seeds/9-seeds
eight_seeds <- play_in_sims %>% filter(play_in == 1 & playoff_seed == 8) %>% 
  select(team, conf, sim_season) %>% 
  mutate(playoff_seed = 8) %>% 
  arrange(conf)
nine_seeds <- play_in_sims %>% filter(play_in == 1 & playoff_seed == 9) %>% 
  select(team, conf, sim_season) %>% 
  mutate(playoff_seed = 9) %>% 
  arrange(conf)

# Series Game Number flag
series_game <- rep(c(1:2), nrow(eight_seeds)) %>% sort()

# Creating fake games 
play_in_games <- rbind(eight_seeds, nine_seeds) %>% 
  cbind(rbind(nine_seeds, eight_seeds)) %>% 
  select(home_team = team, away_team = team1, conf, sim_season, playoff_seed) %>% 
  ungroup() %>%
  dplyr::mutate(game_num = series_game) %>%
  
  # Getting Home team model stats
  left_join(team_data_for_sim %>% 
              
              # Rename vars for home
              dplyr::rename(home_ows = ows, home_dws = dws, home_all_nba = AllNBA), 
            by = c("home_team" = "team")) %>% 
  
  # Joining to get away team team stats
  left_join(team_data_for_sim %>% 
              
              # Rename vars for away
              dplyr::rename(away_ows = ows, away_dws = dws, away_all_nba = AllNBA), 
            by = c("away_team" = "team")) %>% 
  
  
  # Getting differentials for model preds.
  mutate(ows_diff = home_ows - away_ows, 
         dws_diff = home_dws - away_dws, 
         all_nba_diff = home_all_nba - away_all_nba) %>% 

  
  # Getting vars needed
  dplyr::select(home_team, away_team, conf, game_num, sim_season, playoff_seed, ows_diff, dws_diff, all_nba_diff) %>% 
  ungroup()


## Making Predictions on play in games
play_in_model_data <- play_in_games %>% 
  ungroup() %>%
  dplyr::select(ows_diff:all_nba_diff) %>% 
  # mutate_all(scale) %>%
  data.matrix()

play_in_preds <- predict(fit, s = cv_fit$lambda.min, newx = play_in_model_data) %>% as.data.frame() %>% 
  dplyr::rename(pred_score_diff = `1`) %>% 
  mutate(win_prob = predict(model_winprob, ., type = "response")) 

# Sim data with preds 
play_in_final <- play_in_games %>% cbind(play_in_preds)

##### SIM PLAY IN GAMES #####

# random sim
sim_random <- runif(nrow(play_in_final))
  
# Join to play_in_final
sim_df <- play_in_final %>% cbind(sim_random) %>% 
  
  # Simming result based on sim_random
    mutate(sim_win = ifelse(win_prob >= sim_random, 1,0), 
           sim_season = x) %>% 
  dplyr::select(home_team:playoff_seed, sim_win) %>% 
  
  # Getting series winner
  mutate(nine_seed_win_g1 = ifelse(sim_win == 0 & game_num == 1, 1, 0), 
         nine_seed_win_g2 = ifelse(sim_win == 1 & game_num == 2, 1, 0)) %>%
  group_by(conf, sim_season) %>% 
  summarise(nine_seed_wins = sum(nine_seed_win_g1) + sum(nine_seed_win_g2)) %>% 
  
  # Flag for if nine seed won
  mutate(nine_seed_wins = ifelse(nine_seed_wins == 2, 1, 0))

print(x)
return(sim_df)

}

# Run for each simmed season
play_in_output <- lapply(play_in_seasons, play_the_play_in)
play_in_output <- do.call(rbind, play_in_output)

write_rds(play_in_output, "../SunsProject/play_in_simulated.rds")


# Recalculating with play-in winners
play_in_results <- sims %>% group_by(sim_season, conf) %>% 
  arrange(desc(total_wins)) %>% 
  mutate(playoff_seed = rank(-total_wins, ties.method = "random")) %>%
  left_join(play_in_output, by = c("sim_season", "conf")) %>% 
  mutate(made_playoffs = ifelse(playoff_seed <= 7, 1, 0), 
         made_playoffs = ifelse(playoff_seed == 8 & nine_seed_wins == 0, 1, 
                                ifelse(playoff_seed == 9 & nine_seed_wins == 1, 1, made_playoffs))) %>% 
  
  # Getting team playoff probs
  group_by(team, conf) %>% 
  summarise(playin_playoff_prob = round(mean(made_playoffs, na.rm = TRUE), 4))



##################################
########## FINAL STEP ############
##################################

# Join traditional and play in format probs to compare
playoff_comparisons <- traditional_results %>% left_join(play_in_results, by = c("team", "conf")) %>% 
  mutate(prob_diff = playin_playoff_prob - playoff_prob)


# Plotting some stuff
playoff_comparisons %>% ggplot(aes(x = playoff_prob, y = playin_playoff_prob)) + geom_point() + 
  geom_abline() + theme_classic()

#write to rds
write_rds(playoff_comparisons, "../SunsProject/playoff_comparisons.rds")

```


### Playoff Probability Comparison {.tabset}

In this final section, we can see how a play-in implementation effects the playoff odds for each team when compared to a traditional style season. My takeaways from this project are:

- Teams projected to be in the playoffs (1-8 seeds) increase slightly in playoff probability when a play-in is in place. 

- Teams projected to be out of the playoffs (9-15 seeds) decrease slightly in playoff probability when a play-in is in place.

<br/>

#### East Projections

```{r, echo=TRUE}

# Read in playoff comparisons rmd
final_output <- read_rds("../SunsProject/playoff_comparisons.rds") %>% 
  
  # Rounding avg. wins column
  mutate(avg_wins = round(avg_wins,1)) %>%
  
  # Rename columns for Rmd output
  select(Team = team, Conference = conf, `Avg. Wins` = avg_wins, `Traditional Playoff Prob.` = playoff_prob, 
         `Play-in Playoff Prob.` = playin_playoff_prob, Difference = prob_diff)


# Output to Rmd
library(DT)
DT::datatable(final_output %>% filter(Conference == "East") %>% 
                arrange(desc(`Avg. Wins`)), selection = "single",
                  filter = 'top',
                  options = list(sDom  = '<"top"><"bottom">', 
                                  pageLength = 15)) %>% 
      formatPercentage('Traditional Playoff Prob.', digits = 1) %>%
      formatPercentage('Play-in Playoff Prob.', digits = 1) %>%
      formatPercentage('Difference', digits = 2) %>%
      formatStyle('Avg. Wins', background = styleColorBar(final_output$`Avg. Wins`, 'lightblue')) %>% 
      formatStyle('Difference',
                  color = styleInterval(c(-0.0007, 0.0007), c('red', 'gray', 'darkgreen'))
  ) 


```

---

#### West Projections

```{r, echo=TRUE}

# Read in playoff comparisons rmd
final_output <- read_rds("../SunsProject/playoff_comparisons.rds") %>% 
  
  # Rounding avg. wins column
  mutate(avg_wins = round(avg_wins,1)) %>%
  
  # Rename columns for Rmd output
  select(Team = team, Conference = conf, `Avg. Wins` = avg_wins, `Traditional Playoff Prob.` = playoff_prob, 
         `Play-in Playoff Prob.` = playin_playoff_prob, Difference = prob_diff)


# Output to Rmd
library(DT)
DT::datatable(final_output %>% filter(Conference == "West") %>% 
                arrange(desc(`Avg. Wins`)), selection = "single",
                  filter = 'top',
                  options = list(sDom  = '<"top"><"bottom">', 
                                  pageLength = 15)) %>% 
      formatPercentage('Traditional Playoff Prob.', digits = 1) %>%
      formatPercentage('Play-in Playoff Prob.', digits = 1) %>%
      formatPercentage('Difference', digits = 2) %>%
      formatStyle('Avg. Wins', background = styleColorBar(final_output$`Avg. Wins`, 'lightblue')) %>% 
      formatStyle('Difference',
                  color = styleInterval(c(-0.0007, 0.0007), c('red', 'gray', 'darkgreen'))
  ) 


```

---
